\section{Current Electricity}

\subsection{Finding Electric Circuit Components}

\subsubsection*{Learning Objectives}
\begin{itemize}
\item{To identify various components used in electric circuits} 
\item{To find circuit components locally} 
\end{itemize}

\subsubsection*{Background Information}
All of the electrical components studied in secondary school are commonly used devices. These include resistors, capacitors, wires, motors, rheostats, switches, diodes, transistors, transformers, speakers, inductors, bulbs, etc. These components are common enough that they can be found in varying numbers and combinations in any electrical device.  

\subsubsection*{Materials}
broken radio, broken car stereo, broken computer, broken phone charger, broken disc drive, etc.  , pliers, optional soldering iron, screw driver

\subsubsection*{Hazards and Safety}
\begin{itemize}
\item{If you are using a soldering iron, be careful not to touch it or touch someone else with it. It can very quickly cause second degree burns.} 
\item{After breaking and removing components, dispose of any sharp pieces left over.} 
\end{itemize}

\subsubsection*{Preparation Procedure}
\begin{enumerate}
\item{Ask everyone to bring you any broken electrical devices.} 
\item{Go to a radio repair shop in town to find any broken or old components.} 
\end{enumerate}

\subsubsection*{Activity Procedure}
\begin{enumerate}
\item{Open any electrical device.} 
\item{Identify all of the components visible inside the device.} 
\item{Remove as many components as possible from the devices and place them in a container. Pliers can be used to retrieve most things, but sometimes a soldering iron will be necessary to melt the flux holding some components into their boards.} 
\end{enumerate}

\subsubsection*{Results and Conclusion}
Circuit components can be found almost anywhere and can be used to perform many activites.

\subsubsection*{Clean Up Procedure}
\begin{enumerate}
\item{Place all components in a container.} 
\item{Return the remaining useful pieces to their proper places.} 
\item{Dispose of any sharp or otherwise dangerous pieces.} 
\end{enumerate}

\subsubsection*{Discussion Questions}
\begin{enumerate}
\item{What components were you able to retrieve, and from what devices?}
\item{What components were you not able to retrieve, and where might you find them?}
\end{enumerate}

\subsubsection*{Notes}
This should be an ongoing activity for any school. Circuit components are always needed as they are easily destroyed in the laboratory. Keep looking for more things to take apart.  

\subsection{Measuring Emf of a Cell}

\subsubsection*{Learning Objectives}
\begin{itemize}
\item{To identify the terminals of a cell} 
\item{To use a voltmeter or multimeter in parallel} 
\item{To measure the electromotive force of a single cell or battery} 
\end{itemize}

\subsubsection*{Background Information}
Emf is the force used by a cell or battery to move electric current through a circuit. A voltmeter reads the difference between in voltage between two points in a circuit. If the difference in voltage is measured across a cell, the amount of voltage used in the entire circuit, from the positive terminal of the cell to the negative terminal, is being measured. This is the same as measuring the amount of voltage supplied by a cell, or the Emf.  

\subsubsection*{Materials}
Dry cells, speaker wire, multimeter or voltmeter

\subsubsection*{Hazards and Safety}
\begin{itemize}
\item{If you are using a very strong cell (motorcycle battery, etc.  ), do not touch the terminals. The electric current will be too high and will harm you. It is better to use a single dry cell.} 
\end{itemize}

\subsubsection*{Preparation Procedure}
\begin{enumerate}
\item{Collect working dry cells and a multimeter.} 
\end{enumerate}

\subsubsection*{Activity Procedure}
\begin{enumerate}
\item{Set the multimeter to the DCV (direct current voltage) setting.} 
\item{Connect the terminals of the multimeter to the terminals of the battery so that the multimeter displays a voltage level.} 
\item{Adjust the voltage magnitude on the multimeter until the voltage displayed is a clear, readable value.} 
\item{Connect two batteries in series and measure the total Emf}
\item{Connect two batteries in parallel and measure the total Emf}
\end{enumerate}

\subsubsection*{Results and Conclusion}
If a new battery is being used, the voltmeter should display the full Emf of a cell, which is 1.  5 V. If an older battery is being used, the Emf may show a slightly lower value. Students will see that the voltage (Emf) measured across a single cell shows all of the force of the cell without losing any through resistors.  Also, the Emf of batteries in series is more than that of batteries in parallel.

\subsubsection*{Clean Up Procedure}
\begin{enumerate}
\item{Disconnect the multimeter from the dry cells.} 
\item{Turn off the multimeter.} 
\item{Collect all the used materials, storing items that will be used later.} 
\end{enumerate}

\subsubsection*{Discussion Questions}
\begin{enumerate}
\item{What causes the force of the cell?}
\item{What is moving through the circuit?}
\item{Why is a voltmeter connected in parallel across a cell in order to measure Emf?}
\item{What is the difference between the Emf of batteries in series and parallel?}
\item{How would you calculate the Emf of batteries in series?  In parallel?}
\end{enumerate}

\subsubsection*{Notes}
Electromotive force (Emf) is a kind of voltage.  The other kind of voltage is Potential Difference and is used to measure the voltage used by a resistor.  Emf is always the same for a battery; it does not depend on the circuit it is connected to.  Connecting batteries in series simply adds the forces of all the batteries together.
