\chapter{Ionic Theory and Electrolysis}

Atoms form two types of bonds with each other: covalent and electrovalent. In electrovalent bonds, one or more electrons are transferred from a metal atom to a non-metal atoms, resulting in two ions. Many electrovalent compounds dissolve in water to release free ions in solution. Solutions of free ions are called electrolytes. Because these ions can move in solution, electrolytes conduct electricity.

\subsection{Reactivity Series: Displacement Reaction}

Metals can be arranged in order according to their reactivity. The most reactive metals-that is, the ones that are most lihkely to form positive ions-are placed at the top of the list while the least reactive metals-the ones that tend to not form ion-are near the bottom. A metal higher in the reactivity series will displace a metal lower in the series from a solution. 
$$((\mbox {ILLUSTRATION OF REACTIVITY SERIES}))$$
The following activity gives students a chance to demonstrate a displacement reaction in solution. It is best done in small groups so that each student can observe the reaction first-hand.
\subsubsection*{Learning Objectives}
\begin{itemize}
\item{To demonstrate the displacement a metal with a more reactive metal.}
\end{itemize}

\subsubsection*{Materials}
Steel wool*, copper (II) sulphate*, water, and *beaker.

\subsubsection*{Preparation}
\begin{enumerate}
\item{Prepare a copper (II) sulphate solution by dissolving one spoonful of crystals in about 500 mL of water.}
\end{enumerate}

\subsubsection*{Activity Procedure}
\begin{enumerate}
\item {Pour 50 mL of copper (II) sulphate solution into a beaker.}
\item{Examine the steel wool carefully and note its colour. Dip the steel wool into the copper sulphate solution and observe what happens.}
\end{enumerate}

\subsubsection*{Results and Conclusion}
When the steel wool is dipped in the copper sulphate solution, a layer of brown copper metal forms on the surface. This is because iron is higher on the reactivity series than copper, thus iron ions displace copper ions in solution and the copper ions are deposited as copper metal.

\subsubsection*{Clean Up Procedure}
\begin{enumerate}
\item{Collect all the used materials, cleaning and storing items that will be used later.}
\item{Copper sulphate solution can be stored and labelled for future use-it will contain dissolved iron ions. If copper solution is not stored, it should be disposed of in a pit latrine.}
\end{enumerate}

\subsubsection*{Discussion Questions}
\begin{enumerate}
\item{Explain what happens when steel wool is dipped in a solution of copper (II) sulphate.}
\item{Write the chemical equation for the reaction.}
\end{enumerate}

\subsubsection*{Notes}
In the electrochemical series, the more reactive metals displace the less reactive metals. Hence Iron displaces copper in the solution of CuSO4.
Equation.

$$ \mathrm{CuSO}_{4(aq)} + \mathrm{Fe}_{(s)} \longrightarrow \mathrm{FeSO}_{4(aq)} + \mathrm{Cu(s)} $$

\subsection{Electrolytes}

An electrolyte is a substance that forms ions when in solution and thus conducts electricity when dissolved in water. An electrolyte does not conduct electricity when in a solid state because the ions are fixed in a crystal lattice and cannot move.  A strong electrolyte is a substance which is a good conductor of electricity, many ions are formed in solution. A weak electrolyte is a poor conductor of electricity, only few ions are formed in solution. A non-electrolyte is a substance which does not conduct electricity. The following activity is designed for students to classify various substances as electrolytes and non-electrolytes. If the number of battery allows, this activity is best done by students in small groups, however it can also be successfully performed as a class demonstration.

\subsubsection*{Learning Objectives}
\begin{itemize}
\item{To distinguish between strong, weak and non-electrolytes.}
\item{To test the conductivity of electrolytes in solution and in solid form.}
\end{itemize}

\subsubsection*{Materials}
Sodium chloride*, distilled water*, kerosene, copper (II) sulphate*, sugar, 1 M weak acid solution*, 1 M sulphuric acid*, beakers*, stirrer*, 2 carbon electrodes*, connecting wires*, 2-6 V power supply*, 1.5 V bulb*

\subsubsection*{Hazards and Safety}
\begin{itemize}
\item{Concentrated sulphuric acid is corrosive to skin and clothes. Avoid contact with skin and eyes. Neutralize spills with sodium hydrogen carbonate (baking soda).}
\end{itemize}

\subsubsection*{Preparation}
\begin{enumerate}
\item{Set up a circuit with the electrodes, bulb and power source in series. $$\mbox{ILLUSTRATION REQUIRED}$$ Test the circuit by touching the electrodes and confirming that the bulb lights.}
\item{Prepare about 100 mL of approximately 1 M solutions of sulphuric acid and citric or acetic acid.}
\end{enumerate}

\subsubsection*{Activity Procedure}
\begin{enumerate}

\item{Test the conductivity of pure water by putting both electrodes in a beaker of pure water.}
\item{On a piece of paper put a small pile of sodium chloride. Test if the solid sample conducts electricity but putting both dry electrodes in the pile. \textbf{Note:} Make sure the electrodes do not touch each other.}
\item{Pour the sodium chloride into the pure water and stir until dissolved. Instruct the student to now test the salt solution for conductivity.}
\item{Clean the electrodes with distilled water.}
\item{Repeat the test with citric acid crystals, first testing the conductivity of the solid crystals and then adding to pure water and testing the solution.}
\item{Clean the electrodes with distilled water.}
\item{Repeat the test with sugar crystals, first testing the conductivity of the solid crystals and then adding to pure water and testing the solution. Guide students to discuss their observations for the three solutions.}
\item{In one beaker put the 1~M sulphuric acid, in another beaker put 1~M citric acid. Put the electrodes first in the sulphuric acid. Rinse the electrodes and then put them in the citric acid solution. Note the difference in the brightness of the bulb.}
\end{enumerate}

\subsubsection*{Results and Conclusion}
No substance will conduct electricity in the solid state but some of them will conduct in the dissolved state.

The sodium chloride is a strong electrolyte. When electrodes are inserted into salt solution the bulb will light and will be bright.

Citric acid is a weak electrolyte. When electrodes are inserted into citric acid solution, the bulb will light but it will be dimmer than in the sodium chloride solution.

Sugar is a non-electrolyte. The bulb will not light.

The bulb should burn brighter in the sulphuric acid solution than in the citric acid solution even though they are the exact same concentration. This is because the sulphuric acid dissociates completely, you have 1~M solution of hydrogen ions. The citric acid is weak, thus will only partially dissociate-making it a weak electrolyte.

\subsubsection*{Clean Up Procedure}
\begin{enumerate}
\item{Collect all the used materials, cleaning and storing items that will be used later. No special waste disposal is required.}
\end{enumerate}

\subsubsection*{Discussion Questions}
\begin{enumerate}
\item{Classify the substances tested as strong electrolyte, weak electrolyte or non-electrolyte.}
\item{How does this experiment demonstrate the difference between a strong electrolyte and a weak electrolyte?}
\item{Explain using a diagram why solid sodium chloride does not conduct electricity while sodium chloride solution does.}
\end{enumerate}

\subsubsection*{Notes}
Water does not need to be distilled, but prior to the experiment you should confirm that your tap water is not hard enough to conduct electricity.
If time and resources are available, different electrolytes and non electrolytes can be tested such as.
\textbf{Strong electrolytes:} copper (II) chloride, potassium iodide, magnesium sulphate, sodium carbonate
\textbf{Weak electrolytes:} acetic acid (vinegar), lemon juice
\textbf{Non-electrolytes:} ethanol, kerosene

\subsection{Conservation of Energy}
Chemical energy is potential energy stored in the bonds of chemical compounds. Electrical energy is energy from the movement of charge. It is possible to convert from chemical energy to electrical energy and vise versa. This activity give students a change to see the conversion between chemical energy and electrical energy. If materials allow, this activity is great for small groups of students, however if only one ammeter is available the teacher can do it as a demonstration. Care should be taken that each student gets a chance to observe the reaction. 
\subsubsection*{Learning Objectives}
\begin{itemize}
\item{To explain the law of conservation of energy.}
\item{To demonstrate the conversion of chemical energy to electrical energy.}
\end{itemize}

\subsubsection*{Materials}
Copper (II) sulphate, zinc plate*, copper metal*, sand paper, and ammeter.

\subsubsection*{Preparation}
\begin{enumerate}
\item{Prepare a 2~M solution of copper (II) sulphate.}
\item{Clean the solid piece of copper and zinc using steel wool or sand paper.}
\end{enumerate}

\subsubsection*{Activity Procedure}
\begin{enumerate}
\item{Connect the following materials in a series: zinc cathode-connecting wire-ammeter-copper anode. 
$$\mbox{ILLUSTRATION NEEDED}$$ }
\item{Dip the zinc and copper electrodes into the copper (II) sulphate solution.}
\item{Read the current on the ammeter.}
\end{enumerate}

\subsubsection*{Results and Conclusion}
When the circuit is connected a current of 0.06~A is indicated on the ammeter. The current on the ammeter is likely to read 0.06~A but may vary according to the concentration of solution and size of electrodes. The electric current produced indicates that the chemical energy inherent in the electrodes and the electrolyte solution is converted to electrical energy.

\subsubsection*{Clean Up Procedure}
\begin{enumerate}
\item{Collect all the used materials, cleaning and storing items that will be used later.}
\item{Copper sulphate solution can be stored for later use.}
\end{enumerate}

\subsubsection*{Discussion Questions}
\begin{enumerate}
\item{Is it possible to convert electrical energy to chemical energy? How?}
\end{enumerate}

%\subsection{Mechanism of Electrolysis}

\subsection{Electrolysis: Effect of Electrolyte}

Electrolysis is the use of electricity to drive an otherwise unfavourable electrochemical reaction. When electricity is passed through an electrolyte, ions move towards the electrode which attracts them strongest. At each electrode there is an electrochemical reaction which creates a product at that electrode. The nature of ions present in the electrolyte solution affect which products are formed at the cathode and anode.

\subsubsection*{Learning Objectives}
\begin{itemize}
\item{To set up electrolytic cells using different electrolytes.}
\item{To explain how choice of electrolyte can affect the products of electrolysis.}
\end{itemize}

\subsubsection*{Materials}
 copper (II) sulphate*, potassium iodide solution, magnesium sulphate*, beaker*, carbon electrodes*, connection wire, masking tape, 3 to 6V power source*, sand paper

\subsubsection*{Preparation}
\begin{enumerate}
\item{Prepare a solution of copper (II) sulphate by adding approximately one teaspoon of copper (II) sulphate crystals to about 200 mL of water (or reuse the copper sulphate solution from the Electrolysis: Effect of Electrode experiment)}
\item{Prepare about 200 mL of 10\% of potassium iodide solution.}
\item{Prepare a solution of magnesium sulphate by dissolving approximately one teaspoon of crystals in about 200 mL of water.}
\item{Use wire to connect one carbon electrode to either end of the power supply. [electrode diagram]}
\end{enumerate}

\subsubsection*{Activity Procedure}
\begin{enumerate}
\item{In one clear beaker put copper (II) sulphate solution}
\item{Clean the electrodes using sand paper and place them into the beaker of copper sulphate, making sure that the electrodes do not touch each other.  Have students record their observations. After 3 minutes, remove the electrodes and allow students to examine them carefully.}
\item{Clean the electrodes well using water and sand paper. Repeat steps 1 and 2, this time with potassium iodide solution instead of copper sulphate.}
\item{Repeat with magnesium sulphate solution. Make sure to clean the electrodes well each time and record all of your observations.}
\end{enumerate}

\subsubsection*{Results and Conclusion}
When copper sulphate is electrolysed solid brown copper metal will form at the cathode while at the anode oxygen gas will form. This gas can be collected and tested using a glowing splint.
When potassium iodide is electrolysed, hydrogen gas will form at the cathode while at the anode dark iodine solid will form.
When magnesium sulphate is electrolysed, hydrogen gas will form at the cathode while at the anode oxygen gas will form.

\subsubsection*{Clean Up Procedure}
\begin{enumerate}
\item{Collect all the used materials, cleaning and storing items that will be used later. Electrodes should be cleaned well using sand paper and water.}
\item{Copper (II) sulphate solution that is still blue should be kept in a bottle "copper (II) sulphate" to be used in future experiments. Copper (II) sulphate solution that has lost its colour is now a solution of dilute sulphuric acid. Label and store for later use.}
\item{Potassium iodide solution now contains iodine. Ascorbic acid can be used to reduce the iodine back to iodine ions and the solution can be used again.}
\end{enumerate}

\subsubsection*{Discussion Questions}
\begin{enumerate}
\item{Write half reactions for the reactions occurring at the anode and cathode for each of the electrolytes tested.}
\item{What effect did changing the electrolyte have on the products of the electrolysis?}
\item{Explain why changing the electrolyte in this experiment changed the products formed.}
\end{enumerate}

\subsubsection*{Notes}
The products of electrolysis depend on three conditions:
1. Concentration of the electrolyte solution
2. Nature of the electrode (especially anode)
3. Position of the ions in the electrochemical series
DIAGRAM OF ELECTROCHEMICAL SERIES
This experiment examines the third factor. Ions lower in the electrochemical series will be discharged in preference to ions higher in the series. For example, SO42- ions will remain in solution while OH- is dischared. H+ ions will be discharged in preference to K+ ions.
In this experiment because water is present, OH- and H+ ions are available in each solution. The salt dissolved provides competing ions. When copper sulphate solution is present, copper is deposited at the cathode in preference to hydrogen gas because it is lower in reactivity series than hydrogen. On the other hand, when potassium iodide is electrolysed, the potassium remains in solution while H+ ions are reduced at the cathode creating hydrgen gas.
If halogen ions compete with hydroxide ions, the halogen ions will be discharged at the anode if they are in high enough concentration.

\subsection{Electrolysis: Effect of Electrode}

Electrolysis is the use of electricity to drive an otherwise unfavourable electrochemical reaction. When electricity is passed through an electrolyte, ions move towards the electrode which attracts them strongest. At each electrode there is an electrochemical reaction which creates a product at that electrode. The material that the electrode is made of can affect which product is formed at that electrode.

\subsubsection*{Learning Objectives}
\begin{itemize}
\item{To set up an electrolytic cell.}
\item{To explain how choice of electrode can affect the product of electrolysis.}
\end{itemize}

\subsubsection*{Materials}
copper (II) sulphate*, beaker*, carbon electrodes*, copper electrodes*, connection wire, masking tape, 3 to 6V power source*, sand paper

\subsubsection*{Preparation}
\begin{enumerate}
\item{Prepare a solution of copper (II) sulphate by adding approximately one teaspoon of copper (II) sulphate crystals to about 200 mL of water.}
\item{Use wire to connect one electrode to either end of the power supply. [DIAGRAM]}
\item{Use sand paper and water to clean the electrodes.}
\end{enumerate}

\subsubsection*{Activity Procedure}
\begin{enumerate}
\item{In one clear beaker put copper (II) sulphate solution.}
\item{Put both carbon electrodes into the beaker of copper sulphate, making sure that the electrodes do not touch each other.}
\item{Identify the cathode and anode. Record observations about what happens at each electrode. After 3 minutes remove the electrodes from solution and allow students to examine them carefully. Have students record their observations.}
\item{Repeat the experiment using the copper electrodes instead of the carbon electrodes. Guide students to observe the appearance of both electrodes before and after the electrolysis.}
\end{enumerate}

\subsubsection*{Results and Conclusion}
(1) When copper (II) sulphate is electrolysed using the carbon electrodes the blue colour should fade as copper ions are deposited as copper metal on the cathode.  At the anode gas bubbles will be observed forming. The gas produced is oxygen. This can be proven if it is collected in a test tube and tested with a glowing splint (oxygen will relight the glowing splint).
(2) When copper (II) sulphate is electrolysed using copper electrodes the solution remains blue. Copper ions are deposited as metal at the cathode and at the anode reduction of the copper regenerates the ions in solution. After some time, it should be observed that the anode becomes smaller as the copper is oxidized while the anode becomes larger as the copper ions are deposited (reduced). Concentration and pH of the solution should remain unchanged.

\subsubsection*{Clean Up Procedure}
\begin{enumerate}
\item{Collect all the used materials, cleaning and storing items that will be used later. Electrodes should be cleaned with sand paper and water. No special waste disposal is required.}
\item{Copper (II) sulphate solution that is still blue should be kept in a bottle "copper (II) sulphate" to be used in future experiments.}
\item{Copper (II) sulphate solution that has lost its colour is now a solution of dilute sulphuric acid. Label and store for later use.}
\end{enumerate}

\subsubsection*{Discussion Questions}
\begin{enumerate}
\item{Write half reactions for the reactions occurring at the anode and cathode (a) for the carbon electrodes (b) for the copper electrodes.}
\item{What effect did changing the electrodes from carbon to copper have on the products of the electrolysis?}
\item{Explain why changing the electrode in this experiment changed the products formed.}
\item{Explain which products you would expect to see in an experiment with (a) a carbon anode and a copper cathode and (b) a copper anode and a carbon cathode.}
\end{enumerate}

\subsection{Faraday's First Law of Electrolysis}

\subsubsection*{Learning Objectives}
\begin{itemize}
\item{To explain the meaning and significance of Faraday's First Law of Electrolysis.}
\item{To verify Faraday's First Law experimentally.}
\end{itemize}

\subsubsection*{Materials}
magnesium sulphate*, water, 2 burettes of 50 mL capacity, two clean carbon electrodes*, connecting wires*, masking tape, 4-6V power supply*, ammeter, stopwatch*, 1.5 L plastic water bottle, 10 L bucket

\subsubsection*{Preparation}
\begin{enumerate}
\item{Cut the bottom 6 cm from the bottle of the water bottle to form a short beaker.}
\item{Use tape to connect the power supply, ammeter, and 2 carbon electrodes in the following circuit: (ILLUSTRATION)
Leave one of the wires unconnected from the ammeter. During the activity you will connect it.}
\end{enumerate}

\subsubsection*{Activity Procedure}
\begin{enumerate}
\item{Dissolve 5 spoons of magnesium sulphate in about 300 mL water.}
\item{Close the stopcock of each burette and fill both with magnesium sulphate solution.}
\item{Insert one electrode into the top of each burette.}
\item{Add the rest of the magnesium sulphate solution to the beaker.}
\item{Turn the bucket upside-down and place it next to the beaker so that the two containers touch.}
\item{Place your thumb over the open end of one burette. Invert the burette so that the end with the stopcock is pointing up and lower the burette into the beaker. Do not remove your thumb until the open end of the burette it well below the top of the solution in the beaker. The solution in the burette should remain in the burette as long as the stopcock is closed.}
\item{Tape the burette to the bucket of water. Keep the open end under the solution in the beaker! The burette should be pointed almost straight up. (ILLUSTRATION)}
\item{Repeat the previous two steps for the other burette.}
\item{Connect the remaining wire to the ammeter to complete the circuit and immediately start the stopwatch. Bubbles should form at both electrodes and all gas produced should enter into the burettes for collection.}
\item{Record the reading on the ammeter and the volume of gas produced at the cathode after every minute for ten minutes.}
\item{ADD CALCULATIONS IN THIS SECTION}
\end{enumerate}

\subsubsection*{Results and Conclusion}

\subsubsection*{Clean Up Procedure}
\begin{enumerate}
\item{The chemicals used in this experiment do not require any special disposal.}
\end{enumerate}

\subsubsection*{Notes}
If magnesium sulphate is unavailable, dilute sulphuric acid* may be used instead. This is more dangerous, however. If using sulphuric acid, save the final solution - it is still dilute sulphuric acid and may be used for other experiments. While other salts can be used as electrolytes, many will produce other products and are therefore not advised. The calculations for this experiment depend on knowing the amount of electricity passing in the system; therefore the ammeter is required. If an ammeter is unavailable, the teacher should still demonstrate electrolysis, and should still verify Faraday's Second Law.

\subsection{Faraday's Second Law of Electrolysis}

\subsubsection*{Learning Objectives}
\begin{itemize}
\item{SWBAT demonstrate Faraday's second law of electrolysis through the electrolysis of water.}
\end{itemize}

\subsubsection*{Background Information}
We have seen amount of substance deposited or liberated at an electrode depends on the amount of electricity passed through solution. Faraday noted that the quantity also depended on the substance itself. Specifically, the number of moles of a given substance liberated or deposited is inversely proportional to the valence of that substance. This is because the higher the valence, the more electrons are required to liberate it. Hydrogen and oxygen have different valences therefore the electrolysis of water can be used to demonstrate faraday's second law. 
The following activity is useful for students to see for themselves Faraday's Second Law. With two burettes, this activity makes a good demonstration, but if burettes are not available and syringes are used it is best for students to perform the experiments in small groups.

\subsubsection*{Materials}
Magnesium sulphate*, water, two burettes (or two syringes and super glue), carbon electrodes*, connection wire, 3V power source*, beaker

\subsubsection*{Preparation Procedure}
\begin{enumerate}
\item{Prepare a solution of magnesium sulphate by dissolving about 3 spoonfuls of salt into 500 mL of water in a beaker.}
\item{}
\begin{enumerate}
\item{\textbf{If burettes are used:}}
\begin{enumerate}
\item{Connect each electrode to a connecting wire using masking tape.}
\item{Close the stopcock of each burette and fill it with magnesium sulphate solution.} 
\item{Insert one electrode into each burette.}
\item{Place your thumb over the open end of one burette. Invert the burette so that the end with the stopcock is pointing up and lower the burette into the beaker. Do not remove your thumb until the open end of the burette it well below the top of the solution in the beaker. The solution in the burette should remain in the burette as long as the stopcock is closed.}
\item{Tape the burette to the bucket of water. Keep the open end under the solution in the beaker! The burette should be pointed almost straight up. (ILLUSTRATION)}
\item{Repeat the previous two steps for the other burette.}
\item{Carefully open the stopcock slightly to allow air to enter to the 0 mL mark of the burette. It is not important that the water reaches exactly that mark but record the initial volume of air present in each burette.}
\end{enumerate}
\item{\textbf{if burettes are not available, gas collecting electrodes can be constructed using a electrode in a syringe:}}
\begin{enumerate}
\item{Remove the needle from the syringe and use a metal tool to make the hole big enough for the electrode to fit snuggly.}
\item{Insert the syringe so that there is approximately 0.5-1 cm sticking out of the top of the syringe, add super glue to create an air and water tight seal. Connect a wire to each of the electrodes using masking tape.
 (ILLUSTRATION)}
\item{Fill each syringe with magnesium sulphate solution, use a thumb to cover the open end of one syringe. Invert the syringe, making sure that no liquid leaks out and lower the syringe into the beaker containing magnesium sulphate. Do not remove your thumb until the open end of the syringe is below the top of the water. The solution in the syringe should remain in the solution as long as there are no leaks.}
\item{Repeat the previous step for the second syringe.}
\end{enumerate}
\end{enumerate}
\end{enumerate}

\subsubsection*{Activity Procedure}
\begin{enumerate}
\item{Connect the wires in series with the power source. After every minute record the volume of gas collected at each electrode. Repeat this at least 10 times or until the syringe fills.}
\item{Test the gas at each electrode using a glowing splint.}
\end{enumerate}

\subsubsection*{Results and Conclusion}
At the cathode oxygen is reduced to liberate oxygen gas
\ce{2OH-} $\longrightarrow$ \ce{O2} + \ce{2H+} + \ce{4e-}
At the anode, hydrogen is oxidized to liberate hydrogen gas
\ce{2H+}$_{(aq)}$ + \ce{2e-} $\longrightarrow$ \ce{H2}$_{(g)}$
Hydrogen has a valence of 1 while oxygen has a valence of 2. This means that for every 4 moles of electrons passed, one mole of oxygen is liberated while 2 moles of hydrogen are liberated, thus, the volume of hydrogen gas collected should be double the volume of oxygen gas collected.

\subsubsection*{Clean Up Procedure}
\begin{enumerate}
\item{Collect all the used materials, cleaning and storing items that will be used later. No special waste disposal is required.}
\item{If syringes are used, the electrode devices can be stored and used again for later experiments.}
\end{enumerate}

\subsubsection*{Discussion Questions}
\begin{enumerate}
\item{Which gas is produced at the cathode? At the anode?}
\item{Write the electrode half reactions for this electrolysis.}
\item{What volume of each gas was produced? Do these numbers agree with Faraday's Second Law? Explain.}
\end{enumerate}

\subsubsection*{Notes}
In a syringe it might be difficult to know the exact volumes of gases collected. In that case, the experiment is more qualitative than quantitative. The important concept for students to grasp is that there is more hydrogen than oxygen, and it should be clear that there is about 2 times as much hydrogen as oxygen.
Any gas leak in the system will distort results.

\subsection{Electroplating}

Electroplating is the process of coating a metal by depositing another metal through an electrolytic process. Electroplating is used to coat steel materials with other metals such as silver, gold or chromium. Electroplating is popular in automobile manufacture, jewelry making and the making of many materials used in hospitals and in the kitchen.

\subsubsection*{Learning Objectives}
\begin{itemize}
\item{To explain the meaning and significance of electroplating.}
\item{To electroplate a metal with another metal.}
\end{itemize}

\subsubsection*{Materials}
copper (II) sulphate*, iron nail, sand paper, 4-6V  power source*, connection wires, masking tape, beakers*, carbon electrode*

\subsubsection*{Preparation}
\begin{enumerate}
\item{Clean an iron nail thoroughly using sand paper.}
\item{Connect the power source in series with the iron nail and the carbon electrode with the iron nail as the cathode (connected to the + end of the battery) and the carbon electrode as the anode (connected to the - end of the battery).}
\item{Prepare a 2~M solution of copper (II) sulphate and acidify it by adding some dilute sulphuric acid.}
\end{enumerate}

\subsubsection*{Activity Procedure}
\begin{enumerate}
\item{Insert the nail and the carbon electrode into the acidified copper (II) sulphate solution.}
\item{Observe and record the changes on both the electrodes after 30 minutes.}
\end{enumerate}

\subsubsection*{Results and Conclusion}
The nail will become coated in a thin layer of brown copper metals-we have electroplated iron with copper. At the carbon electrode (anode) oxygen gas will be formed.

\subsubsection*{Clean Up Procedure}
\begin{enumerate}
\item{The copper sulphate can be kept and labelled as "acidified copper sulphate" for future use.}
\item{The nail may be cleaned using sand paper.}
\item{Collect all the used materials, cleaning and storing items that will be used later.}
\end{enumerate}

\subsubsection*{Notes}
If the copper coat does not stick well to the nail, it may be necessary to soak it in dilute sulphuric acid and then cleaned with sand paper or steel wool prior to the experiment.